---
title: Data transformations
author: "Lucas Veras"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    number_sections: true
    toc_float: 
      collapsed: true
    highlight: tango
    df_print: paged
bibliography: bibliography.bib
csl: nlm.csl
---

```{r source, eval = TRUE, echo = FALSE, include = FALSE}
library(here)
source(here("code", "01_clean_data.R"))
```

# Basic transformations

To run the meta-analysis of longitudinal data, the values needed are the mean and standard deviation of the change score (pre-post data), either in absolute or percentage values. Nevertheless, the studies report not only the mean and standard deviation, but a series of other central tendency and dispersion measures, that ought to be transformed prior to running a meta-analysis.

In this section it will be covered the transformations from other central tendency measures to the mean and from other dispersion measures to the standard deviation.

## Central tendency measures

Among the studies selected to the meta-analysis the reported central tendency measures were the mean and the median.

### From median to mean

$$ \bar{X} = \frac{Q_1 + \tilde{X} + Q_3}{3} $$

Where $\bar{X}$ is the sample mean, $\tilde{X}$ is the median and $Q_1$ and $Q_3$ are the 1st and 3rd quartiles, respectively.

This transformation was proposed in a paper by Wan and colleagues [@Wan2014].

## Dispersion measures

The dispersion measures reported in the studies were the standard error of the mean, the interquartile range, and the confidence interval.

### From standard error to standard deviation

$$ s = SE\cdot\sqrt{n} $$

Where $s$ is the standard deviation, $SE$ is the standard error and $n$ is the sample size.

### From interquartile range to standard deviation

$$ s = \frac{Q_3 - Q_1}{2\phi^{-1}(\frac{0.75n-0.125}{n+0.25})} $$

Where $s$ is the standard deviation, $Q_1$ and $Q_3$ are the 1st and 3rd quartiles, respectively, $n$ is the sample size and $\phi{-1}$ is the inverse of the quantile function of the normal distribution.

This transformation was also proposed by Wan et al. [@Wan2014]

### From the confidence interval to the standard deviation

The confidence interval is defined as:

$$ CI = \bar{X} \pm t_{df}\frac{s}{\sqrt{n}} $$

Where $CI$ is the confidence interval, $s$ is the standard deviation, $n$ is the sample size and $t_{df}$ is the quantile of the $t$ distribution with $n - 1$ degrees of freedom ($df$).

The "margin of error" $t_{df}\frac{s}{\sqrt{n}}$ is added and subtracted from the mean value to calculate the confidence interval, and therefore, it corresponds to half the width of the confidence interval. So:

$$ \frac{CI_{upper} - CI_{lower}}{2} = t_{df}\frac{s}{\sqrt{n}} $$
$$ s = \frac{(CI_{upper} - CI_{lower}) \cdot \sqrt{n}}{2t_{df}} $$

# Transformations to percentage change

To run a meta-analysis with the percentage change, its mean and standard deviation are needed.

To calculate the mean percentage change based on the absolute values:

$$ \bar{X}_{\%\ change} = \frac{\bar{X}_{post} - \bar{X}_{pre}}{\bar{X}_{pre}} \cdot 100 $$

Where $\bar{X}_{\%\ change}$ is the mean percentage change and $\bar{X}_{pre}$ and $\bar{X}_{post}$ are the means of the pre and post scores, respectively.

And to calculate the standard deviation of the mean percentage change:

$$ s_{\%\ change} = \frac{s_{mean\ change}}{s_{pre}} $$
$$ s_{mean\ change} = \sqrt{s^2_{pre} + s^2_{post} - 2 \cdot r \cdot s_{pre} \cdot s_{post}} $$

Where $s_{\%\ change}$ is the standard deviation of the mean percentage change, $s_{mean\ change}$ is the standard deviation of the absolute mean change, $s_{pre}$ and $s_{post}$ are the standard deviation of the pre and post scores, respectively, and $r$ is the correlation coefficient between the pre and post scores.

These transformations were done accordingly to the procedures described by Shams-White et al. [@Shams-White2017]

Below is a list of the variables, per study and time after surgery, that needed this transformation.

```{r abs_to_pct, echo = FALSE}
data_percentage_change_cmp %>%
  select(
    Study = id,
    `Time after surgery` = time_after_surgery,
    Variable = outcome
  ) %>%
  filter(
    Variable %!in% c("body_mass", "BMI", "fat_mass", "lean_mass", "age")
  )
```

# References
