---
title: Data transformations
author: "Lucas Veras"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    number_sections: true
    toc_float: 
      collapsed: true
    highlight: tango
    df_print: paged
editor_options: 
  chunk_output_type: console
---

```{r source, eval = TRUE, echo = FALSE, include = FALSE}
library(here)
source(here("code", "01_clean_data.R"))
```

# Basic transformations

To run the meta-analysis of longitudinal data, the values needed are the mean and standard deviation of the change score (pre-post data), either in absolute or percentage values. Nevertheless, the studies report not only the mean and standard deviation, but a series of other central tendency and dispersion measures, that ought to be transformed prior to running a meta-analysis.

In this section it will be covered the transformations from other central tendency measures to the mean and from other dispersion measures to the standard deviation.

## Central tendency measures

Among the studies selected to the meta-analysis the reported central tendency measures were the mean and the median.

### From median to mean

$$ \bar{X} = \frac{Q_1 + \tilde{X} + Q_3}{3} $$

Where $\bar{X}$ is the sample mean, $\tilde{X}$ is the median and $Q_1$ and $Q_3$ are the 1st and 3rd quartiles, respectively.

This transformation was proposed in the paper Wan et al. BMC Medical Research Methodology 2014, 14:135.

## Dispersion measures

The dispersion measures reported in the studies were the standard error of the mean, the interquartile range, and the confidence interval.

### From standard error to standard deviation

$$ s = SE\cdot\sqrt{n} $$

Where $s$ is the standard deviation, $SE$ is the standard error and $n$ is the sample size.

### From interquartile range to standard deviation

$$ s = \frac{Q_3 - Q_1}{2\phi^{-1}(\frac{0.75n-0.125}{n+0.25})} $$

Where $s$ is the standard deviation, $Q_1$ and $Q_3$ are the 1st and 3rd quartiles, respectively, $n$ is the sample size and $\phi{-1}$ is the inverse of the quantile function of the normal distribution.

This transformation was proposed in the paper Wan et al. BMC Medical Research Methodology 2014, 14:135.

### From the confidence interval to the standard deviation

The confidence interval is defined as:

$$ CI = \bar{X} \pm t_{df}\frac{s}{\sqrt{n}} $$

Where $CI$ is the confidence interval, $s$ is the standard deviation, $n$ is the sample size and $t_{df}$ is the quantile of the $t$ distribution with $n - 1$ degrees of freedom ($df$).

The "margin of error" $t_{df}\frac{s}{\sqrt{n}}$ is added and subtracted from the mean value to calculate the confidence interval, and therefore, it corresponds to half the width of the confidence interval. So:

$$ \frac{CI_{upper} - CI_{lower}}{2} = t_{df}\frac{s}{\sqrt{n}} $$
$$ s = \frac{(CI_{upper} - CI_{lower}) \cdot \sqrt{n}}{2t_{df}} $$

# Transformations to percentage change

To run a meta-analysis with the percentage change, its mean and standard deviation are needed.

To calculate the mean percentage change based on the absolute values:

$$ \bar{X}_{\%\ change} = \frac{\bar{X}_{post} - \bar{X}_{pre}}{\bar{X}_{pre}} \cdot 100 $$

Where $\bar{X}_{\%\ change}$ is the mean percentage change and $\bar{X}_{pre}$ and $\bar{X}_{post}$ are the means of the pre and post scores, respectively.

And to calculate the standard deviation of the mean percentage change:

$$ s_{\%\ change} = \frac{s_{mean\ change}}{s_{pre}} $$
$$ s_{mean\ change} = \sqrt{s^2_{pre} + s^2_{post} - 2 \cdot r \cdot s_{pre} \cdot s_{post}} $$

Where $s_{\%\ change}$ is the standard deviation of the mean percentage change, $s_{mean\ change}$ is the standard deviation of the absolute mean change, $s_{pre}$ and $s_{post}$ are the standard deviation of the pre and post scores, respectively, and $r$ is the correlation coefficient between the pre and post scores.

Below is a list of the variables, per study and time after surgery, that needed this transformation.

```{r abs_to_pct, echo = FALSE}
data_percentage_change_cmp %>%
  select(
    Study = id,
    `Time after surgery` = time_after_surgery,
    Variable = outcome
  )
```

# Transformations to absolute change

As with the percentage change, to use the absolute change, its mean and standard deviation is needed.

To compute the mean absolute change, simply:

$$ \bar{X}_{mean\ change} = \bar{X}_{post} - \bar{X}_{pre} $$

Where $\bar{X}_{mean\ change}$, $\bar{X}_{pre}$ and $\bar{X}_{post}$ are the mean absolute change and the means of the pre and post scores, respectively.

The computation of the standard deviation of the mean absolute change was already defined in the previous section.

Below is a list of the variables, per study and time after surgery, that needed this transformation.

```{r abs_to_chn, echo = FALSE}
abs_to_chn <- data %>%
  filter(
    !is.na(mean) | !is.na(sd) |
    !is.na(median) | !is.na(iqr_lower) | !is.na(iqr_upper)
  ) %>%
  select(
    Study = id,
    `Time after surgery` = time_after_surgery,
    Variable = outcome
  )
abs_to_chn
```
